A Study on Common Bug Fix Patterns in Rust

### 2022-10-30 methodology.tex
methodology comments; probably need to move things around

@@

The general idea for automatically finding the pervasive code patterns within a list of programs is to parse the programs and, for us, to embed the important parts of the derived AST in fixed-size datapoints. Then, using a clustering algorithm, we can group these datapoints in different clusters. Work in this vein makes the assumption that each cluster corresponds to one class of change pattern~\cite{hanam2016discovering,campos2019discovering,yang2022mining}.

Our pipeline for finding classes of pervasive patterns in Rust programs also follows this idea. We decided to implement our pipeline completely in Python. To make this work, we need a way to get an AST as a Python dictionary. Syn\footnote{\url{https://crates.io/crates/syn}} is a Rust crate which aims to serve Rust procedural macros. However, it includes a parser which is suitable for our purposes; we simply had to write a preprocessor to transform Syn AST output into Python dictionaries.

To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs, encoded in two Python dictionaries. Tree diff algorithms can compute the differences between two arbitrary trees. When the trees are programs' abstract syntax trees, we call their diff an ASTDiff. We use dictdiffer to obtain ASTDiffs.

An ASTDiff may include an arbitrary number of semantic changes, although a best practice is to include only one semantic change in a commit. Because that best practice is not universally followed, we are interested in finding the most important change within an ASTDiff, and embedding that change in our datapoints. Sections~\ref{sec:path_extraction}~and~\ref{sec:weighting_scheme} provide a detailed explanation of how we select the most important semantic information, which then allowed us to obtain clusters that contained similar datapoints.

Our target repositories were the top 18 most-starred open source Rust projects on GitHub. We mined (using Pydriller) all the bug related commits and ran them through our pipeline. We will provide more details on the aforementioned process later. Then, we applied the DBSCAN clustering algorithm~\cite{ester1996density} on the resulting datapoints. In Section~\ref{sec:clustering_data} we justify our choice of DBSCAN for clustering, and we explain how we tuned it to improve cluster quality.

@@

@@

As mentioned above, we use Syn for parsing Rust. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related elements (BCE) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \verb+BCE+ elements can be found within Appendix~\ref{sec:appendix}.

@@

\end{figure}

@@

The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change (terminals and nonterminals). Our tool looks for changes within each root-level scope. That is, our assumption is that a change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool would detect both continuous and non-continuous line changes within one scope.

Looking at the tree encoded in the content part of each diff, we realized that each path to a leaf represents one contributing element to the change. A Depth First Search allows us to collect all paths, and we return only the paths that we store in our datapoints (i.e. touching $\mathtt{NT} \cup \mathtt{BCE}$).

Figure~\ref{fig:extraction} illustrates path extraction on a simple change. The change replaces \verb+tail+ with \verb+&tail+ and \verb+head+ with \verb+head.clone()+. That is, after the change, the second argument passes a borrow of \verb+tail+ rather than sending the ownership of \verb+tail+ to the callee; and, in the third argument, instead of sending \verb+head+ to the callee, it sends a clone of it. A DFS on this tree yields three different paths, which we show in red, green, and purple. A path represents a sequence of involved non terminals. We found it crucial to record the order of nodes in the path, as this gives us the freedom to later record the order in our datapoint embedding.

\paragraph{Code Embedding.} A key question is how to transform these paths to a fixed size datapoint. First, we need to define a fixed set of columns. We need fixed-size datapoints, as they are a requirement for both centroid-based and density-based well-known clustering algorithms~\cite{xu2005survey}. The number of columns would depend on the amount of information we want to embed in the datapoints.

The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an embedding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to embed all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCE}$ as our dimensions, which yields an infeasible $O(n^d)$ dimensions ($n=|\mathtt{NT} \cup \mathtt{BCE}|$ and $d$ is the depth of nested structures); this could potentially be reduced somewhat, but is still impractical. While we can implement a specific workaround to distinguish nested ifs (and we have done so), the general point about the large dimensionality of fully order-sensitive embedding still holds.

@@

In our methodology, we propose a novel representation. Similar to the first approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCE}$ we collected from Syn. However, to simulate order-sensitiveness, we carried out two auxiliary steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

\subsection{\label{sec:mining_repositories}Mining Repositories}

@@

Now that we have set up our code analysis pipeline, we can start mining the Rust repositories. We target the top 18 most-starred Rust projects on GitHub, at the time of data collection. Table~\ref{table:repos} summarizes our benchmarks. Because we are interested in changes, we computed the average code churn of the files within the projects (The code churn is computed using process metrics provided by Pydriller). denoland/deno and swc-project/swc are the projects with largest number of LOCs and highest average code churn, respectively. Unsurprisingly, we captured a lot of instances from these two projects in our final clusters.

@@

\subsection{\label{sec:clustering_data}Clustering Data}

Having collected our datapoints, our next task is to cluster them so that we can categorize bug patterns. Like~\cite{hanam2016discovering}, we use the DBSCAN clustering algorithm for two main reasons. Firstly, in contrast to k-means, it does not require the number of clusters in advance as an input to the algorithm. Secondly, it is a density-based clustering method, which means that it detects arbitrarily shaped clusters; centroid-based methods like k-means can't detect such clusters. We tried k-means and the more powerful fuzzy variant Fuzzy C-means~\cite{dunn1973fuzzy} and did not get useful clusters.

DBSCAN takes two tuneable parameters. The first parameter $\epsilon$ indicates a radius within which a point is a core point, a border point, or an outlier. The second parameter $Z$ is the minimum number of points per cluster. We ran DBSCAN using different combinations of these two parameters. In Figure~\ref{fig:clustering}, we show nine different experiments, with their respective parameter values. Each subplot specifies, for both $D_g$ and $D_b$, the number of clustered points, noise points, and the number of clusters.

% TODO fix this caption
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{clusters.pdf}
\caption{\label{fig:clustering} Clustering results with nine different combinations of parameters $\epsilon$ and $Z$, respectively separated in columns and rows. In each subplot, the two blue bars in the left respectively indicate the number of clustered and noise datapoints (the number is shown above each bar) after applying clustering on $D_g$ (denoted with $G_c$ and $G_n$ on the x axis). The two orange bars in the right show similar data but for the $D_{b}$ database. In each subplot's legend, $C_G$ and $C_B$ show the number of clusters obtained from $D_g$ and $D_b$, respectively}
\end{figure}

As a general rule, lower $Z$ would allow more groups of close points to be considered as clusters, resulting in a higher number of clusters. A large $\epsilon$ would widen the search radius, allowing more points to fall in the clusters, which results in a reduction of the number of noise points. Such a choice might put points from different patterns inside the same cluster, resulting in a ineffective clustering. Alternatively, a small $\epsilon$ would tighten the ring of search, possibly separating clusters which essentially manifest the same code changes. Also, if $\epsilon$ is really small, the clustering algorithm might not detect some clusters at all, resulting in a reduction of the number of clusters.

@@

That is why we carried out a manual analysis of clusters. In the manual analysis, we randomly pick 50 clusters; from each, cluster we randomly choose 10 datapoints. The person analyzing the cluster then looks at the code and specifies the clusters that contained more than five datapoints showing similar patterns. For the manual analysis, we used $\epsilon=0.0001, Z=5$ and $\epsilon=0.001, Z=5$, which were the best parameters for $D_g$ and $D_b$, resulting in 577 and 102 clusters, respectively. As a final step, we exclude clusters that contained datapoints from fewer than 3 projects. That is because we wanted our clusters to contain cross-project (e.g. potentially generalizable) patterns as much as possible.

@@

Similar to~\cite{yang2022mining}, following Cotroneo et al.~\cite{cotroneo2019analyzing} definitions, we divide the clusters in three different groups:

@@

In this work, we were only interested in the first two groups and ignored the clusters for refactoring changes. Also, we prefer reporting clusters that manifest concrete patterns. An abstract pattern as opposed to a concrete pattern can account for many changes in the programs. For instance, we deem the pattern `Changing a clone of a variable to a borrowing of it in a function argument' more concrete than `Adding a new statement to a function's body':

@@

In our manual analysis, we randomly picked 50 datapoints of each cluster. If the cluster had fewer than 50 datapoints, we would analyze all of the datapoints in the cluster. Next, we read the code and the bug reports related to all the datapoints (if they existed). If we judged that the members of a cluster fit into a pattern, we wrote a natural language description of that cluster. Considering the descriptions, we linked similar clusters together as possible candidates for merging. We considered these candidates and carried out appropriate ones. Finally, we selected 19 of the remaining clusters---the ones that manifested concrete changes. We discuss about our findings in Sections~\ref{sec:common_patterns}~and~\ref{sec:bc_patterns}.

>>>>

\todo{introductory para saying that we're going to describe our methodology}

@@

Our approach for automatically finding pervasive code patterns aims to find clusters of similar code patterns, using existing implementations of appropriate clustering algorithms. Work in this vein generally makes the assumption that clusters corresponds to classes of change patterns~\cite{hanam2016discovering,campos2019discovering,yang2022mining}. We thus want to associate code changes with datapoints; one of our contributions is a fixed-size embedding which highlights the important parts of a code change.

We now briefly describe our methodology, providing full details in later subsections. Our pipeline for finding classes of pervasive patterns in Rust programs parses program versions, computes differences between them, embeds them in fixed-size datapoints, and then clusters these points.

\todo[inline]{I think we only need to talk about this next bit in a subsection. It's an implementation detail, not fundamental to our approach: ``We decided to implement our pipeline completely in Python. To make this work, we need a way to get an AST as a Python dictionary.''}

\todo[inline]{I wonder if this should really go here, or if there should be a shorter version.}
To compute the contents of a change, we need two code revisions: the revision before the commit, and the revision after it. After parsing these two revisions, we will have two ASTs. Tree diff algorithms can compute the differences between two arbitrary trees. When the trees are programs' abstract syntax trees, we call their diff an ASTDiff. 
% , encoded in two Python dictionaries; We use dictdiffer to obtain ASTDiffs.
\todo[inline]{Maybe this next paragraph is what you want at the top of 2.1. Then in the next next paragraph you put a shorter description of the pipeline, in a sentence before ``we mined'', and you reference 2.1 there. Anyway think about organization and try to have something that makes sense.}
An ASTDiff may include an arbitrary number of semantic changes, although a best practice is to include only one semantic change in a commit. Because that best practice is not universally followed, we are interested in finding the most important change within an ASTDiff, and embedding that change in our datapoints. Sections~\ref{sec:path_extraction}~and~\ref{sec:weighting_scheme} provide a detailed explanation of how we select the most important semantic information, which then allowed us to obtain clusters that contained similar datapoints.

Our target repositories were the top 18 most-starred open source Rust projects on GitHub\todo{as of what date?}. We mined (using Pydriller) all the bug related commits and ran them through our pipeline; Section~\ref{sec:mining} describes our mining methodology in depth. Then, we applied the DBSCAN clustering algorithm~\cite{ester1996density} on the resulting datapoints. Section~\ref{sec:clustering_data} describes clustering in detail, justifies our choice of DBSCAN for clustering, and explains how we tuned it to improve cluster quality.

@@

\todo[inline]{should always be text between subsection and subsubsection. This is the per-change workflow, right?}

@@

Syn\footnote{\url{https://crates.io/crates/syn}} is a Rust crate built for the Rust procedural macro implementation. However, it includes a parser which is suitable for our purposes; we simply had to write a preprocessor to transform Syn AST output into Python dictionaries. Syn handles all of Rust. Each of our datapoints specifies whether any Syn non terminal (\texttt{NT}) is present or absent, i.e. containing one dimension per non terminal. However, we are also interested in change patterns that involve the borrow checker (BC), so we also add a dimension reflecting the presence of a set of BC-related elements (BCE) that we identified, e.g. \texttt{clone}, \texttt{Rc}, \texttt{Box}. The full list of \verb+BCE+ elements can be found in Appendix~\ref{sec:appendix}.

@@

\end{figure}
% yay tikz

@@

The second part identifies the context of the modification, that is, the path from the root of the tree to the subtree in which the modification has occured. The third part is the content of the change---terminals and nonterminals. Our tool\todo{do we have a name? It's important to have a name} looks for changes within each root-level scope. That is, our assumption is that each change pattern occurs within one root-level scope. In this work, we chose to not handle patterns that involve changes in multiple functions or multiple files. However, our tool detects both continuous and non-continuous line changes within one scope.

Looking at the tree encoded in the content part of each diff, we realized that each path to a leaf represents one element contributing to the change. A Depth First Search allows us to collect all paths, and we return only the paths that we store in our datapoints (i.e. touching $\mathtt{NT} \cup \mathtt{BCE}$).
\todo{if you mathtt BCE here, you should do it everywhere, i.e. in the paragraph at the start of 2.1.1.}

Figure~\ref{fig:extraction} illustrates path extraction on a simple change. The change replaces \verb+tail+ with \verb+&tail+ and \verb+head+ with \verb+head.clone()+. That is, after the change, the second argument passes a borrow of \verb+tail+ rather than sending the ownership of \verb+tail+ to the callee; and, in the third argument, instead of sending \verb+head+ to the callee, it sends a clone of it. A DFS on this tree yields three different paths, which we show in red, green, and purple. A path represents a sequence of involved non terminals. We found it crucial to record the order of nodes in the path, as this allows us to record that order in our datapoint embedding.

\paragraph{Code Embedding.} A key question is how to transform these paths to a fixed size datapoint. First, we need to define a fixed set of columns. We need fixed-size datapoints, as they are a requirement for well-known centroid-based and density-based clustering algorithms~\cite{xu2005survey}. The number of columns controls the amount of information we can embed in the datapoints.

The most naive approach would be to only report the observed non terminals within the diff. That yields a fully order-insensitive representation. In such an embedding, for instance, there is no difference between two nested if statements and two if statements beside each other. On the other hand, theoretically, for full order-sensitiveness, we would need to embed all possible combinations of items from $\mathtt{NT} \cup \mathtt{BCE}$ as our dimensions, which yields an infeasible $O(n^d)$ dimensions ($n=|\mathtt{NT} \cup \mathtt{BCE}|$ and $d$ is the depth of nested structures); this could potentially be reduced somewhat, but is still impractical. While we can implement a specific workaround to distinguish nested ifs (and we have done so), the general point about the large dimensionality of fully order-sensitive embedding still holds.
\todo[inline]{Worse than that, isn't $d$ unbounded?}

@@

We propose a novel representation. Similar to the first\todo{which? the naive one?} approach, we order-insensitively define the columns of our dataset as the set of items $\mathtt{NT} \cup \mathtt{BCE}$ we collected from Syn. However, to simulate order-sensitiveness, we add two more steps. First, we record the number of occurrences of items within the paths at each scope. The reason behind this decision is because a different order of program elements yields a different number of occurrences of underlying dimension items \todo{is there an example?}. Second, we semi-automatically design a weighting scheme to prioritize non terminals that we think are more salient for Rust bugs.

@@

\subsection{\label{sec:mining_repositories}Mining Repositories}
\label{sec:mining}

@@

Now that we have set up our code analysis pipeline, we can start mining the Rust repositories. We target the top 18 most-starred Rust projects on GitHub, at the time of data collection. Table~\ref{table:repos} summarizes our benchmarks. Because we are interested in changes, we computed (using process metrics provided by our mining library) the average code churn of the files within the projects. denoland/deno and swc-project/swc are the projects with largest number of LOCs and highest average code churn, respectively. Unsurprisingly, we captured a lot of instances from these two projects in our final clusters.

@@

\subsection{\label{sec:clustering_data}Clustering Data}
\label{sec:clustering}

Having collected our datapoints, our next task is to cluster them so that we can categorize bug patterns. Like~\cite{hanam2016discovering}, we use the DBSCAN clustering algorithm for two main reasons. Firstly, in contrast to k-means, it does not require the number of clusters in advance as an input to the algorithm. Secondly, it is a density-based clustering method, which means that it detects arbitrarily shaped clusters; centroid-based methods like k-means can't detect such clusters. We did try k-means and its more powerful fuzzy variant Fuzzy C-means~\cite{dunn1973fuzzy} and did not get useful clusters.

DBSCAN takes two tuneable parameters. The first parameter $\epsilon$ indicates a radius distinguishing core points, border points, and outliers. The second parameter $Z$ is the minimum number of points per cluster. We ran DBSCAN using different combinations of these two parameters. In Figure~\ref{fig:clustering}, we show nine different experiments, with their respective parameter values. Each subplot specifies, for both $D_g$ and $D_b$, the number of clustered points, noise points, and the number of clusters.

% TODO fix this caption
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{clusters.pdf}
\caption{\label{fig:clustering} Clustering results with nine different combinations of parameters $\epsilon$ and $Z$, respectively separated in columns and rows. In each subplot, the two blue bars in the left respectively indicate the number of clustered ($G_c$) and noise ($G_n)$ datapoints (the number is shown above each bar) after applying clustering on general database $D_g$. The two orange bars in the right show similar data but for the borrow-checker $D_{b}$ database. In each subplot's legend, $C_G$ and $C_B$ show the number of clusters obtained from $D_g$ and $D_b$, respectively}
\end{figure}

\todo{figure caption: why the capital G and B in the subscript?}

As a general rule, lower $Z$ would allow more groups of nearby points to be considered to be clusters, resulting in a higher number of clusters. A large $\epsilon$ would widen the search radius, allowing more points to fall in the clusters, which results in a reduction of the number of noise points. However, such a choice might put points from different patterns inside the same cluster, resulting in ineffective clustering. Alternatively, a small $\epsilon$ would tighten the ring of search, possibly separating clusters which essentially manifest the same code changes. Also, if $\epsilon$ is really small, the clustering algorithm might not detect some clusters at all, resulting in a reduction of the number of clusters.

@@

\todo{that is why what? huh? be more explicit and repeat yourself a bit}
That is why we carried out a manual analysis of clusters. In the manual analysis, we randomly pick 50 clusters; from each cluster we randomly choose 10 datapoints. The person analyzing the cluster \todo{who? be more specific about how we did this} then looks at the code and specifies the clusters that contained more than five datapoints showing similar patterns. For the manual analysis, we used $\epsilon=0.0001, Z=5$ and $\epsilon=0.001, Z=5$, which were the best parameters for $D_g$ and $D_b$, resulting in 577 and 102 clusters, respectively. As a final step, we exclude clusters that contained datapoints from fewer than 3 projects. That is because we wanted our clusters to contain cross-project (e.g. potentially generalizable) patterns as much as possible.

@@

Like~\cite{yang2022mining}, and following the definitions of Cotroneo et al.~\cite{cotroneo2019analyzing}, we divide the clusters in three different groups:

@@

\todo{can we say more-specific rather than concrete?}
In this work, we were only interested in the first two groups and ignored the clusters for refactoring changes. Also, we prefer reporting clusters that manifest concrete patterns. An abstract pattern as opposed to a concrete pattern can account for many changes in the programs. For instance, we deem the pattern `Changing a clone of a variable to a borrowing of it in a function argument' more concrete than `Adding a new statement to a function's body':

@@

In our manual analysis, we randomly picked 50 datapoints of each cluster. If the cluster had fewer than 50 datapoints, we would analyze all of the datapoints in the cluster. Next, we read the code and the bug reports related to all the datapoints (if they existed). If we judged that the members of a cluster fit into a pattern, we wrote a natural language description of that cluster. Considering the descriptions, we linked similar clusters together as possible candidates for merging. We considered these candidates and carried out appropriate ones. Finally, we selected 19 of the remaining clusters---the ones that manifested concrete changes. Sections~\ref{sec:common_patterns}~and~\ref{sec:bc_patterns} present our clusters.
